---
title: "Peak_width_distribution_liftover"
output: html_notebook
---

 Janis

```{r}
library(ggplot2)
library(dplyr)
library(readr)

# === Define paths ===
hg38_dir <- "cuttag_dir/peak_calling"
lifted_dir <- "cuttag_dir/reciprocal_liftover/"
setwd("cuttag_dir/reciprocal_liftover/")

# === Cell types to include ===
cell_types <- c("FOXA2", "OLIG2", "NEUN")

# === Function to read BED and compute widths ===
process_bed <- function(file_path, cell_type, species) {
  bed <- read_tsv(file_path, col_names = FALSE, col_types = cols(.default = "c"))
  
  # Ensure minimum required columns
  if (ncol(bed) < 4) {
    warning(paste("Skipping file due to insufficient columns:", file_path))
    return(NULL)
  }
  
  bed <- bed %>%
  filter(!grepl("chrX|chrY|random", X1, ignore.case = TRUE)) %>%
  mutate(
    Chr = X1,
    Start = as.numeric(X2),
    End = as.numeric(X3),
    Peak_ID = X4,
    Peak_Width = End - Start,
    Sample = paste0(cell_type, "_", species),
    Cell_Type = cell_type,
    Species = species
  )

  
  return(bed[, c("Chr", "Start", "End", "Peak_ID", "Sample", "Cell_Type", "Species", "Peak_Width")])
}

# === Read all peaks ===
all_peak_data <- bind_rows(lapply(cell_types, function(ct) {
  hg38_file <- file.path(hg38_dir, paste0(ct, "_confidence_peaks.sorted.bed"))
  hg19_file <- file.path(lifted_dir, paste0(ct, "_final_human_peaks_noXYrandom.bed"))
  
  bind_rows(
    process_bed(hg38_file, ct, "hg38"),
    process_bed(hg19_file, ct, "hg19_lifted")
  )
}))

# === Format and transform ===
all_peak_data <- all_peak_data %>%
  filter(!is.na(Peak_Width) & Peak_Width > 0) %>%
  mutate(
    log10_Peak_Width = log10(Peak_Width),
    Sample = factor(Sample, levels = unique(Sample))
  )


# === Cell type colors ===
cell_type_colors <- c("NEUN" = "#009E73", "FOXA2" = "#785594", "OLIG2" = "#0072B2")

```

# === Plot ===
pdf("Lifted_vs_hg38_Peak_Width_Boxplot_log10.pdf", width = 5, height = 3)

ggplot(all_peak_data, aes(x = Sample, y = log10_Peak_Width, fill = Cell_Type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = cell_type_colors) +
  labs(
    title = "Log10 Peak Widths: Original hg38 vs Lifted hg19 Peaks",
    x = "Sample (Cell type + Species)",
    y = "log10(Peak Width) (bp)"
  ) +
  coord_flip() +  # 🔄 Flip axes
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold")) 


dev.off()

```{r}
library(dplyr)
library(ggplot2)

# Subset hg38 and hg19 peaks separately
hg38_peaks <- all_peak_data %>% filter(Species == "hg38")
hg19_peaks <- all_peak_data %>% filter(Species == "hg19_lifted")

# Join by Peak_ID and Cell_Type to ensure alignment
matched_peaks <- inner_join(hg38_peaks, hg19_peaks, 
                            by = c("Peak_ID", "Cell_Type"), 
                            suffix = c("_hg38", "_hg19"))

# Optionally log10-transform widths
matched_peaks <- matched_peaks %>%
  mutate(
    log10_hg38 = log10(Peak_Width_hg38),
    log10_hg19 = log10(Peak_Width_hg19)
  )

# One model per cell type
lm_results <- matched_peaks %>%
  group_by(Cell_Type) %>%
  do({
    model <- lm(log10_hg19 ~ log10_hg38, data = .)
    data.frame(
      R2 = summary(model)$r.squared
    )
})

print(lm_results)


library(ggplot2)
library(ggpmisc)  # for stat_poly_eq

p <- ggplot(matched_peaks, aes(x = log10_hg38, y = log10_hg19, color = Cell_Type)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~Cell_Type) +
  stat_poly_eq(
    aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")),
    formula = y ~ x,
    parse = TRUE,
    label.x = "left",
    label.y = "top",
    size = 5,
    color = "black"
  ) +
  labs(
    title = "Correlation of Peak Widths (Hg38 vs Lifted Hg19)",
    x = "log10(Peak Width) - hg38",
    y = "log10(Peak Width) - Lifted hg19"
  ) +
  scale_color_manual(values = cell_type_colors) +
  theme_minimal()

ggsave("peak_correlation_hg19_lift_hg38.png", p, width = 15, height = 5, dpi = 300)


```

```{r}
# === Count peaks per sample ===
peak_counts <- all_peak_data %>%
  group_by(Cell_Type, Species) %>%
  summarize(Total_Peaks = n(), .groups = "drop") %>%
  mutate(Sample = paste0(Cell_Type, "_", Species))

# Reorder for consistent x-axis
peak_counts$Sample <- factor(peak_counts$Sample, levels = unique(peak_counts$Sample))

pdf("Lifted_vs_hg38_Peak_numbers_Bars.pdf", width = 5, height = 3)
ggplot(peak_counts, aes(x = Sample, y = Total_Peaks, fill = Cell_Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = cell_type_colors) +
  labs(
    title = "Number of Peaks: hg38 vs Lifted hg19",
    x = "Sample (Cell type + Species)",
    y = "Total Number of Peaks"
  ) +
  theme_minimal(base_size = 13) +
  coord_flip() +  # 🔄 Flip axes
  theme(
    legend.position = "right",
    axis.text.y = element_text(hjust = 1)
  )
dev.off()


```